---
title: "R-Ladies Philly Shiny Dashboard Exercises (Advanced Topics)"
author: "Anastasia Lucas"
date: "1/2/2021"
runtime: shiny
output: html_document
---
### Set up

We will need two additional packages, ```dashboardthemes``` and ```plotly```, to complete the exercises in this document.

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(shiny)
library(shinydashboard)
library(dashboardthemes)
library(plotly)
```

### Read in our data

```{r data}
# Choose from c("arianagrande", "beyonce", "drake", "queen", "taylorswift", "thebeatles")
artist_name <- "thebeatles"
df <- read.delim(paste0("data/", artist_name, "_lyrics.txt"))
```

### Adding error messages using validate
We set our numeric input so the user could select between 3 and 20 words. But if the user types in the input box, it will override our defaults. Let's say we want to restrict our app so the user will get an error message if they try to display fewer than 2 or more than 50 words.\
We can do this using ```validate```, which tests a condition and returns an error if the test fails, and ```need```, which consists of an expression that evaluates to TRUE or FALSE and an error message.\
Hint: the code to create a validation function is below. We would then need to place our validation function ```v()``` within each ```render*()``` function that we want it to affect.

```{hint 2}

v <- reactive({
  validate(
    need(input$select_topn < 2 | input$select_topn > 50, "Please select a number between 3 and 50")
  )
})

```

```{r exercise 7, eval=FALSE}

ui <- dashboardPage(
  dashboardHeader(title = "Song Lyrics"),
  dashboardSidebar(  
    numericInput(inputId = "select_topn", 
              label = "Select the number of words to display", 
              value = 10, min = 3, max = 20, step=1)
  ),
  dashboardBody(    
    fluidPage(
      fluidRow(
        column(# width should be between 1 and 12
               width=6,
               box(plotOutput("plot1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value
                   width=NULL)
              ),
        column(# width should be between 1 and 12
               width=6,
               box(dataTableOutput("table1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value
                   width=NULL)
              )
      )
    )
  )
)

server <- function(input, output) { 
  # Create our validation function
  v <- ____________
})
  
  output$plot1 <- renderPlot({
    # Validate for bar chart
    ____________
    
    df %>% 
      group_by(word) %>% 
      summarise(occurrences=sum(n)) %>% 
      arrange(desc(occurrences)) %>% 
      top_n(input$select_topn) %>%
      ungroup() %>%
      ggplot(aes(x=word, 
                 y=occurrences)) +
      geom_col() +
      ylab("count") +
      coord_flip() +
      theme_minimal() +
      scale_fill_gradient(high = "#f6a97a", low="#ca3c97") +
      ggtitle(paste("Top", input$select_topn, "frequently used words")) + 
      geom_blank()
  })
  
  output$table1 <- renderDataTable({
    # Validate for table
    ____________
    # We could also replace v() with a different validate function like we defined above
    # validate(
    #  need(input$select_topn >= 2 & input$select_topn <= 50, "Please select a number between 3 and 50")
    # )
    df %>% 
      group_by(word) %>% 
      summarise(occurrences=sum(n)) %>% 
      arrange(desc(occurrences)) %>% 
      top_n(input$select_topn) %>%
      ungroup()
  })
}

shinyApp(ui, server)

```

Bonus: See if you can restrict the plot to display between 3 and 50 words, but the table to display between 1 and 250 words.

```{r bonus 1, eval=FALSE}

ui <- dashboardPage(
  dashboardHeader(title = "Song Lyrics"),
  dashboardSidebar(  
    numericInput(inputId = "select_topn", 
              label = "Select the number of words to display", 
              value = 10, min = 3, max = 20, step=1)
  ),
  dashboardBody(   
    fluidPage(
      fluidRow(
        column(# width should be between 1 and 12
               width=6,
               box(plotOutput("plot1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value
                   width=NULL)
              ),
        column(# width should be between 1 and 12
               width=6,
               box(dataTableOutput("table1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value
                   width=NULL)
              )
      )
    )
  )
)

server <- function(input, output) { 
  # Create our validation function
  output$plot1 <- renderPlot({
    # Validate for bar chart
    validate(
      ____________
    )
    
    df %>% 
      group_by(word) %>% 
      summarise(occurrences=sum(n)) %>% 
      arrange(desc(occurrences)) %>% 
      top_n(input$select_topn) %>%
      ungroup() %>%
      ggplot(aes(x=word, 
                 y=occurrences)) +
      geom_col() +
      ylab("count") +
      coord_flip() +
      theme_minimal() +
      scale_fill_gradient(high = "#f6a97a", low="#ca3c97") +
      ggtitle(paste("Top", input$select_topn, "frequently used words")) + 
      geom_blank()
  })
  
  output$table1 <- renderDataTable({
    # Validate for table
    validate(
      ____________
    )
    
    df %>% 
      group_by(word) %>% 
      summarise(occurrences=sum(n)) %>% 
      arrange(desc(occurrences)) %>% 
      top_n(input$select_topn) %>%
      ungroup()
  })
}

shinyApp(ui, server)

```


### Okay now let's get to some fun stuff
Try to change the theme of your dashboard using themes from the ```dashboardthemes``` package. We'll want to add a call to ```dashboardthemes::shinyDashboardThemes()``` in our ```dashboardBody```. Run ```?dashboardthemes::shinyDashboardThemes``` to see a list and description of the available themes.\
These can be helpful if you don't want to change the CSS manually. (As a non-CSS/HTML expert, I usually don't!)

```{r exercise 8, eval=FALSE}

ui <- dashboardPage(
  dashboardHeader(title = "Song Lyrics"),
  dashboardSidebar(  
    numericInput(inputId = "select_topn", 
              label = "Select the number of words to display", 
              value = 10, min = 3, max = 20, step=1)
  ),
  dashboardBody(  
    # Change theme here
   dashboardthemes::shinyDashboardThemes(
     theme = ____________
   ), 
    
    fluidPage(
      fluidRow(
        column(# width should be between 1 and 12
               width=6,
               box(plotOutput("plot1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value
                   width=NULL)
              ),
        column(# width should be between 1 and 12
               width=6,
               box(dataTableOutput("table1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value
                   width=NULL)
              )
      )
    )
  )
)

server <- function(input, output) { 
  # Create our validation function
  v <- reactive({
    validate(
      need(input$select_topn >= 2 & input$select_topn <= 50, "Please select a number between 3 and 50")
    )
})
  
  output$plot1 <- renderPlot({
    # Validate for bar chart
    v()
    
    df %>% 
      group_by(word) %>% 
      summarise(occurrences=sum(n)) %>% 
      arrange(desc(occurrences)) %>% 
      top_n(input$select_topn) %>%
      ungroup() %>%
      ggplot(aes(x=word, 
                 y=occurrences)) +
      geom_col() +
      ylab("count") +
      coord_flip() +
      theme_minimal() +
      scale_fill_gradient(high = "#f6a97a", low="#ca3c97") +
      ggtitle(paste("Top", input$select_topn, "frequently used words")) + 
      geom_blank()
  })
  
  output$table1 <- renderDataTable({
    # Validate for table
    v()
    
    df %>% 
      group_by(word) %>% 
      summarise(occurrences=sum(n)) %>% 
      arrange(desc(occurrences)) %>% 
      top_n(input$select_topn) %>%
      ungroup()
  })
}

shinyApp(ui, server)

```

This package is still in active development and as a result does have some limitations as far as customization goes. Luckily they do provide some nice DIY functions to create your own themes and logos. 

### Let's try making our plot interactive using Plotly
We could use base Plotly or use some of the click/hover arguments in native Shiny's ```plotOutput()``` to create our interactive charts. But, to make it even easier on ourselves, we can also wrap our ggplot in ```plotly::ggplotly()``` to easily convert our ggplot object into a plotly object. The code below shows an example of ggplot object -> plotly object conversion.\
Try to incorporate this code into your dashbaord. 

```{r hint 3}
    
   plotly::ggplotly( 
     df %>% 
       group_by(word) %>% 
       summarise(occurrences=sum(n)) %>% 
       arrange(desc(occurrences)) %>% 
       top_n(10) %>%
       ungroup() %>%
       ggplot(aes(x=word, 
                  y=occurrences)) +
       geom_col() +
       ylab("count") +
       coord_flip() +
       theme_minimal() +
       scale_fill_gradient(high = "#f6a97a", low="#ca3c97") +
       ggtitle(paste("Top", 10, "frequently used words")) + 
       geom_blank()
   )

```

Hint: We will also will need to tell shinydashboard that our resulting plot should be a Plotly object instead of a static chart. This means we will need to change ```plotOutput``` to ```plotly::plotlyOutput``` and ```renderPlot``` to ```plotly::renderPlotly```.

```{r exercise 9, eval=FALSE}

ui <- dashboardPage(
  dashboardHeader(title = "Song Lyrics"),
  dashboardSidebar(  
    numericInput(inputId = "select_topn", 
              label = "Select the number of words to display", 
              value = 10, min = 3, max = 20, step=1)
  ),
  dashboardBody(  
    # Change theme here
   dashboardthemes::shinyDashboardThemes(
     theme = "onenote"
   ), 
    
    fluidPage(
      fluidRow(
        column(# width should be between 1 and 12
               width=6,
               box(____________("plot1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value
                   width=NULL)
              ),
        column(# width should be between 1 and 12
               width=6,
               box(dataTableOutput("table1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value
                   width=NULL)
              )
      )
    )
  )
)

server <- function(input, output) { 
  # Create our validation function
  v <- reactive({
    validate(
      need(input$select_topn >= 2 & input$select_topn <= 50, "Please select a number between 3 and 50")
    )
})
  
  output$plot1 <- ____________({
    # Validate for bar chart
    v()
   ____________ 
     df %>% 
       group_by(word) %>% 
       summarise(occurrences=sum(n)) %>% 
       arrange(desc(occurrences)) %>% 
       top_n(input$select_topn) %>%
       ungroup() %>%
       ggplot(aes(x=word, 
                  y=occurrences)) +
       geom_col() +
       ylab("count") +
       coord_flip() +
       theme_minimal() +
       scale_fill_gradient(high = "#f6a97a", low="#ca3c97") +
       ggtitle(paste("Top", input$select_topn, "frequently used words")) + 
       geom_blank()
   ____________
  })
  
  output$table1 <- renderDataTable({
    # Validate for table
    v()
    
    df %>% 
      group_by(word) %>% 
      summarise(occurrences=sum(n)) %>% 
      arrange(desc(occurrences)) %>% 
      top_n(input$select_topn) %>%
      ungroup()
  })
}

shinyApp(ui, server)

```

### Lastly, let's try to make our app more efficient
We are working with a pretty small dataset with only a few thousand lines, but code efficiency and performance become even more important as we scale up. For example, in our code we are summarizing the data each time the user makes a new selection. It would be a better idea for us to summarize the data outside of our app.\
Remember we can only summarise up until the point we need our reactive input value!

```{r summarize}
df_sum <- df %>% 
           group_by(word) %>% 
           summarise(occurrences=sum(n)) %>% 
           arrange(desc(occurrences))
# Now we have saved df_sum in our environment and it is ready to use
```

And now we can replace ```df``` in our dashboard with ```df_sum```. If we were making a standalone app, we could save ```df_sum``` to a file and read that file in instead.\
We still need to grab the ```top_n``` values dependent on the user selection. To make our code even cleaner, since our table and bar chart both depend on the same user input value, we really only need to filter once. We can do this outside of the ```render*()``` functions, but we will still need to create a reactive function for our dataframe.\
Hint: Here is some example code for reactively filtering data.

```{r hint 4}
  df_filtered <- reactive({
    df_sum %>%
      top_n(input$select_topn)
  }) 
```

```{r exercise 10, eval=FALSE}

ui <- dashboardPage(
  dashboardHeader(title = "Song Lyrics"),
  dashboardSidebar(  
    numericInput(inputId = "select_topn", 
              label = "Select the number of words to display", 
              value = 10, min = 3, max = 20, step=1)
  ),
  dashboardBody(  
    # Change theme here
   dashboardthemes::shinyDashboardThemes(
     theme = "onenote"
   ), 
    
    fluidPage(
      fluidRow(
        column(# width should be between 1 and 12
               width=6,
               box(plotly::plotlyOutput("plot1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value,
                   width=NULL)
              ),
        column(# width should be between 1 and 12
               width=6,
               box(dataTableOutput("table1"), 
                   title="Most Common Words Across Albums",
                   # For column based layouts, we can set box width to NULL
                   # This overrides the default value
                   width=NULL)
              )
      )
    )
  )
)

server <- function(input, output) { 
  # Create our validation function
  v <- reactive({
    validate(
      need(input$select_topn >= 2 & input$select_topn <= 50, "Please select a number between 3 and 50")
    )
})
  
  df_filtered <- ____________
  
  output$plot1 <- plotly::renderPlotly({
    # Validate for bar chart
    v()
    plotly::ggplotly( 
      ____________ %>% 
        ggplot(aes(x=word, 
                   y=occurrences)) +
        geom_col() +
        ylab("count") +
        coord_flip() +
        theme_minimal() +
        scale_fill_gradient(high = "#f6a97a", low="#ca3c97") +
        ggtitle(paste("Top", input$select_topn, "frequently used words")) + 
        geom_blank()
   )
  })
  
  output$table1 <- renderDataTable({
    # Validate for table
    v()
    ____________
  })
}

shinyApp(ui, server)

```

We could compare the performance of this code to our previous version by using the ```profvis``` package or RStudio's Profile menu (which I think also uses profvis...).\
The difference may be negligable here, but it could have a big impact if we are trying to filter millions of rows.

### Other things to try
1. We could add additional control widgets and plots. For example, it might be useful for the user to be able to find the top n words per album or display the songs with the most unique words. 
2. As with Shiny, you can provide additional customization to your dashbaord and apps with cascading style sheets (CSS) since both are built from HTML. You can read more about using CSS with Shiny and shinydashboard at https://shiny.rstudio.com/articles/css.html and https://rstudio.github.io/shinydashboard/appearance.html#css respectively.\
3. You can also include a variety of icons https://rstudio.github.io/shinydashboard/appearance.html#icons or change the appearance of your boxex by assigning colors https://rstudio.github.io/shinydashboard/appearance.html#statuses-and-colors or making them collapsible. Try replacing lines 526-530 with the code below to see how the different attributes look. 

```{status}

  box(plotly::plotlyOutput("plot1"), 
      title = "Most Common Words Across Albums",
      status = "primary",
      collapsible = TRUE,
      solidHeader = TRUE,
      width=NULL)

```

4. We could also make some improvements to the plots themselves. We might take a look at Trang's December data visualization workshop to get some ideas!
5. Anything else you can think of that will help the user get the maximum amount of information out of your dashboard.
